### 2021-03-09 sections/1-introduction.tex
wrote outline

Cite CHESS~\cite{ishaq2019clustered}

>>>>

Growth in size and types of datasets:
\begin{itemize}
    \item SDSS APOGEE2 data. Near-infrared spectra of approximately 430,000 stars in approximately 9,000 dimensions. $R=22,500 = \frac{\lambda}{\Delta \lambda}$ for $\lambda \in [1.51, 1.70) \mu m$
    \item SDSS MaNGA hexagon shaped images. 10,000 galaxies with 2,700 deg$^2$, resolution $R=2,000$ for $\lambda \in [360, 1000) nm$
    \item Silva 18S ribosomal DNA sequences of approximately 2.25 million genomes with aligned length of 50,000 letters. TODO: Look into other variants in the dataset particularly NR99.
    \item Datasets from ANN-Benchmarks suite
\end{itemize}

Recent approaches in dealing with exponential growth of data:
\begin{itemize}
    \item LSH
    \item clever indexing e.g. FM Index.
    \item Entropy-scaling search~\cite{yu2015entropy} and CHESS~\cite{ishaq2019clustered}.
\end{itemize}

This paper focuses on:
\begin{itemize}
    \item Improvements and extensions of CHESS.
    \item Accelerated $\rho$-nearest and k-nearest neighbors search.
    \item Entropy-scaling (?? is it though?) compression using the hierarchical clustering.
    \item Proof for how cluster-radii, and thus compression deltas, fall as we descend along the tree.
    \item Information theoretic bounds on compression performance. We compress along the dimensionality of data rather than the cardinality of data.
    \item A variety of distance functions including euclidean, manhattan, cosine, jaccard, wasserstein, hamming, levenshtein, needleman-wunch, and any user-defined function.
\end{itemize}

Comparisons for:
\begin{itemize}
    \item Search time vs FALCONN and other in ann-benchmarks suite.
    \item Compression performance of CAKES+gzip vs just gzip.
\end{itemize}

### 2021-03-09 sections/2-methods.tex
wrote outline

\subsection{Proofs}
\label{subserc:methods:proofs}

@@

We take a dataset to be any set $X$ with $n$ instances in $\mathcal{D}$ embedding dimensions where the data follow a low-dimensional manifold with fractal dimension $d$ where $d \ll \mathcal{D}$.

We take a distance function to be $f: X \times X \mapsto \mathbb{R}$ with the following properties:

\begin{itemize}
    \item $f(x, y) = 0 \ iff \ x = y$, identity
    \item $f(x, y) = f(y, x) \ \forall x, y \in X$, symmetry
    \item $f(x - z, y - z) = f(x, y) \ \forall x, y, z \in X$, invariance under translation 
\end{itemize}

We show that the radii of clusters are guaranteed to decrease after every $d$ partitions (at most).

Assume that the data follow a $d$-dimensional distribution embedded in $\mathcal{D}$-dimensional space.
We can describe this distribution choosing some set of $d$ mutually orthagonal axes.
Let $2r$ be the maximum distance between any pair of points along any axis from any choice of mutually orthagonal axes for our $d$-dimensional distribution.
The cluster containing all these points will have a radius of $r$.
In the worst case, the distribution is a $d$-sphere and $2r$ is the maximum distance along every axis from every choice of axes.

The partition algorithm will choose the maximally distant pair of points and, thus, the axis defined by those points as the axis for partitioning the points into two child clusters.
For points in either child cluster, the maximum distance between any pair of points along that axis will have been reduced to $r$ or less.
If $d > 1$ then the child clusters will have a radius $r_{child}$ such that $r_{child} \leq r \sqrt{2}$.
By partitioning the cluster in this way, we have taken one axis and reduced the maximum distance along that axis from $2r$ to $r$.
However, as per the distribution of data as discussed earlier, there may be other axes along which the maximum pairwise distance is still $2r$.
We, thus, recursively partition each child cluster.

With each recursive application of Partition, we consume one axis along which the maximum distance was $2r$ and reduce the maximum distance along that axis to $r$ in the child clusters.
After $d$ recursive applications of partition, we will have exhausted the $d$ axes with the maximum distance of $2r$.
After these $d$ partitions, the points in each child cluster will have a maximum pairwise distance of $r$ along any axis.
Thus the radii of those child clusters will be at most $\frac{r}{2}$.

>>>>

Talk about manifold hypothesis.

\subsection{Definitions}
\label{subsec:methods:definitions}

\subsubsection{Dataset}
\label{subsubsec:methods:definitions:dataset}

General datasets with high dimensionality and high cardinality.
We take a dataset to be any set $X$ with $n$ instances in $\mathcal{D}$ embedding dimensions where the data follow a low-dimensional manifold with fractal dimension $d$ where $d \ll \mathcal{D}$.

\subsubsection{Distance Functions}
\label{subsubsec:methods:definitions:distance-functions}

General distance function which deterministically maps a pair of instances to a non-negative real number:
\begin{itemize}
    \item $dist: \mathbb{X} \times \mathbb{X} \mapsto \mathbb{R}^+$
    \item $dist(x, y) = 0 \Leftrightarrow x = y$, identity
    \item $dist(x, y) = dist(y, x) \ \forall x, y \in X$, symmetry
    \item $dist(x - z, y - z) = dist(x, y) \ \forall x, y, z \in X$, invariance under translation, not valid for cosine distance or when subtraction is not defined.
\end{itemize}

\subsubsection{Local Fractal Dimension}
\label{subsubsec:methods:definitions:local-fractal-dimension}

LFD is a discrete approximation to Hausdorff Dimension:

\begin{gather}
    \log_2\bigg(\frac{|B_D(q, r_1)|}{|B_D(q, r_2)|}\bigg)
    \label{fractal-dimension}
\end{gather}

where $B_D(q,r)$ is the set of instances contained in a ball on the dataset $D$ of radius $r$ centered on an instance $q$.
We approximate the local fractal dimension by computing it for a radius $r_1$ and a smaller radius $r_2=\frac{r_1}{2}$.

\subsubsection{Cluster}
\label{subsubsec:methods:definitions:cluster}

A \textit{cluster} is a set of instances with a \textit{center} (the geometric median) and a \textit{radius} (the maximum distance of any instance in the cluster from the cluster-center).
Each cluster is either a leaf-cluster or has two child-clusters in the same way as nodes do in a binary tree.

\subsubsection{Metric Entropy}
\label{subsubsec:methods:definitions:metric-entropy}

Same as in CHESS.

\subsection{Proofs}
\label{subsec:methods:proofs}

@@

We show that the radii of clusters are guaranteed to decrease after, at most, every $d$ recursive partitions.

Assume that the data follow a $d$-dimensional distribution embedded in $\mathcal{D}$-dimensional space.
We can describe this distribution by choosing some set of $d$ mutually orthagonal axes.
Let $2r$ be the maximum distance between any pair of points along any axis from any choice of mutually orthagonal axes for our $d$-dimensional distribution.
The cluster containing these points will have a radius of $r$.
In the worst case, the distribution is a $d$-sphere and $2r$ is the maximum distance along every axis from every choice of axes.

The partition algorithm will choose the maximally distant pair of points and, thus, the axis defined by those points for partitioning the cluster into two child clusters.
For points in either child cluster, the maximum distance between any pair of points along that axis will have been reduced to $r$ or less.
If $d > 1$ then the child clusters will have a radius $r_{child}$ such that $r_{child} \leq r \sqrt{2}$.
By partitioning the cluster in this way, we have taken one axis and reduced the maximum distance along that axis from $2r$ to $r$.
However, as per the distribution of data as discussed earlier, there may be other axes along which the maximum pairwise distance is still $2r$.
We, thus, recursively partition each child cluster.

With each recursive application of Partition, we consume one axis along which the maximum distance was $2r$ and reduce the maximum distance along that axis to $r$ in the child clusters.
After $d$ recursive applications of partition, we will have exhausted the $d$ axes with the maximum distance of $2r$.
After these $d$ partitions, the points in each child cluster will have a maximum pairwise distance of $r$ along any axis.
Thus the radii of those child clusters will be at most $\frac{r}{\sqrt{2}}$.

\subsubsection{Information-Theoretic properties of Compression}
\label{subsubsec:methods:proofs:information-theoretic-properties-of-compression}

TODO, TODO, TODO!

\subsection{Algorithms}
\label{subsec:methods:algorithms}

\subsubsection{Partition}
\label{subsubsec:methods:algorithms:partition}

Partition algorithm for building the tree.
Lift this from the CHAODA paper.

% latex does not seem to like math symbols in this heading
\subsubsection{$\rho$-Nearest Neighbors Search}
\label{subsubsec:methods:algorithms:rnn-search}

Modified binary search.
Lift this from CHESS paper.

\subsection{k-Nearest Neighbors Search}
\label{subsubsec:methods:algorithms:knn-search}

Repeated $\rho$-nearest neighbors search with modified radius with each iteration guided by lfd of query ball.
Use from CHESS paper as a starting point.

\subsubsection{Compression}
\label{subsubsec:methods:algorithms:compression}

Encode and Decode algorithms.
These will be specific for each dataset and/or distance function.

\subsection{Datasets and Distance Functions}
\label{subsec:methods:datasets-and-distance-functions}

Provide details on the datasets we use for benchmarks.

Add t-SNE or UMAP embeddings of datasets. For example~\ref{fig:discussion:umap-annthyroid-euclidean}

\begin{figure}[ht!]
    \centering
    \includegraphics[width=2.5in]{images/umaps/annthyroid-euclidean-umap2d.png}
    \caption{UMAP embedding of Annthyroid data with the Euclidean distance function.}
    \label{fig:discussion:umap-annthyroid-euclidean}
\end{figure}

\subsubsection{APOGEE-2}
\label{subsec:methods:datasets:apogee-2}

euclidean, cosine, wasserstein-1d

\dots

\subsubsection{MaNGA}
\label{subsec:methods:datasets:manga}

euclidean, wasserstein-2d

\dots

\subsubsection{Silva 18S and variants}
\label{subsec:methods:datasets:silva-18s}

hamming, levenshtein

\dots

\subsubsection{ANN-Benchmarks suite}
\label{subsec:methods:datasets:ann-benchmarks-suite}

deep1b: cosine

fashion-mnist: euclidean

gist: euclidean

glove: cosine

kosarak: jaccard

mnist: euclidean

nytimes: cosine

sift: euclidean

last.fm: cosine

\dots

\subsection{Hardware}
\label{subsec:methods:hardware}

Specs of Ark, and maybe M1 MacBook Air and Dell XPS.

### 2021-03-09 sections/4-discussion.tex
wrote outline

Include figures~\ref{fig:discussion:umap-annthyroid-euclidean}

\begin{figure}[ht!]
    \centering
    \includegraphics[width=2.5in]{images/umaps/annthyroid-euclidean-umap2d.png}
    \caption{UMAP embedding of Annthyroid data with the Euclidean distance function.}
    \label{fig:discussion:umap-annthyroid-euclidean}
\end{figure}

>>>>

CLAM-CAKES.
This is most effective when datasets exhibit low metric entropy and low local fractal dimension.

Unlike LSH, this is extensible to any user provided distance function.

Exactness of search and metric vs non-metric.

\subsection{Future Works}
\label{subsec:results:future-works}

GPU-acceleration for the more computationally intensive distance functions like wasserstein.

Live-updates as new points are added to the dataset.
For example, build tree with a subsample of the full data and then add the rest of the instances to simulate live-updates.

More distance functions. For example, maximal-common-subgraph for molecular structures.

TODO: Provide link to source code on github.
